---
title: "Catch comparison paired tests"
author: "Jason Flower"
date: "19/06/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(knitr)
library(vegan) #for diversity comparison

source("../functions/testing_2samples.R")
```

```{r data import and tidying, message=FALSE}
#full dataset (each fish)
traps <- read_csv("../data/traps_data.csv", col_types = cols_only(
  TrapID = col_character(),
  Surveyors = col_character(),
  Length_cm = col_double(),
  LengthType_FL_TL = col_character(),
  Weight_Grams = col_double(),
  sciname = col_character(),
  Date_YMD = col_date(format = ""),
  Exp_or_Cont = col_character(),
  design = col_character(),
  Deep_Shallow = col_character(),
  days_since_last_haul = col_double(),
  common = col_character(),
  family = col_character(),
  N_coord = col_double(),
  W_coord = col_double(),
  location_exposure = col_character(),
  depth_m = col_double(),
  zero_or_1_fish = col_double()
))

#same dataset, zero fish hauls excluded
traps_no_zeros <- traps %>% 
  filter(zero_or_1_fish != 0)

#to get only paired trap data, summarise to get mean length for each trap, spread the data, and then use na.omit to drop any traps that don't have a mean length value for both Control and Experimental. Gather the data back together
traps_paired <- traps %>% 
  group_by(Date_YMD, TrapID, Exp_or_Cont) %>%
  summarise(Length_mean = mean(Length_cm)) %>%
  #filter(Length_mean != 0) %>%
  mutate(TrapNo = str_extract(TrapID, "\\d+")) %>%
  ungroup() %>%
  select(-TrapID) %>% #need to remove this otherwise spread doesn't work as needed
  spread(Exp_or_Cont, Length_mean) %>%
  na.omit() %>%
  gather(key = "Exp_or_Cont", value = "Length_mean", c("Control", "Experimental")) %>%
  #rebuild TrapID code
  mutate(TrapID = str_c(str_sub(Exp_or_Cont, 1,1), TrapNo)) %>%
  select(-TrapNo) %>%
  distinct(TrapID, Date_YMD) %>%
  inner_join(traps)

#make a dataset excluding the traps hauls with no fish
traps_paired_no_zeros <- traps %>% 
  filter(common != "No fish") %>%
  group_by(Date_YMD, TrapID, Exp_or_Cont) %>%
  summarise(Length_mean = mean(Length_cm)) %>%
  #filter(Length_mean != 0) %>%
  mutate(TrapNo = str_extract(TrapID, "\\d+")) %>%
  ungroup() %>%
  select(-TrapID) %>% #need to remove this otherwise spread doesn't work as needed
  spread(Exp_or_Cont, Length_mean) %>%
  na.omit() %>%
  gather(key = "Exp_or_Cont", value = "Length_mean", c("Control", "Experimental")) %>%
  #rebuild TrapID code
  mutate(TrapID = str_c(str_sub(Exp_or_Cont, 1,1), TrapNo)) %>%
  select(-TrapNo) %>%
  distinct(TrapID, Date_YMD) %>%
  inner_join(traps)

#import data summarized at the haul level
trap_hauls <- read_csv("../data/traps_hauls.csv", col_types = cols_only(
  TrapID = col_character(),
  Date_YMD = col_date(format = ""),
  Exp_or_Cont = col_character(),
  design = col_character(),
  days_since_last_haul = col_double(),
  depth_m = col_double(),
  N_coord = col_double(),
  W_coord = col_double(),
  location_exposure = col_character(),
  mean_length_cm = col_double(),
  total_biomass_kg = col_double(),
  no_fish = col_double(),
  no_species = col_double()
))

#summarize data for hauls
trap_hauls_no_zero <- traps_no_zeros %>% 
  group_by(TrapID, Date_YMD, Exp_or_Cont) %>% 
  summarise(mean_length_cm = mean(Length_cm), total_biomass_kg = sum(Weight_Grams)/1000, no_fish = sum(zero_or_1_fish), no_species = n_distinct(sciname), depth_m = mean(depth_m), days_since_last_haul = mean(days_since_last_haul)) %>%
  mutate(no_species = case_when(no_fish == 0 ~ 0,
                                TRUE ~ as.numeric(no_species))) %>% 
  ungroup()

trap_hauls_paired <- traps_paired %>% 
  group_by(TrapID, Date_YMD, Exp_or_Cont) %>% 
  summarise(mean_length_cm = mean(Length_cm), total_biomass_kg = sum(Weight_Grams)/1000, no_fish = sum(zero_or_1_fish), no_species = n_distinct(sciname), depth_m = mean(depth_m), days_since_last_haul = mean(days_since_last_haul)) %>%
  #need this otherwise traps hauls with no fish are counted as having 1 species
  mutate(no_species = case_when(no_fish == 0 ~ 0,
                                TRUE ~ as.numeric(no_species))) %>% 
  ungroup()

trap_hauls_paired_no_zeros <- traps_paired_no_zeros %>% 
  group_by(TrapID, Date_YMD, Exp_or_Cont) %>% 
  summarise(mean_length_cm = mean(Length_cm), total_biomass_kg = sum(Weight_Grams)/1000, no_fish = sum(zero_or_1_fish), no_species = n_distinct(sciname), depth_m = mean(depth_m), days_since_last_haul = mean(days_since_last_haul)) %>%
  #need this otherwise traps hauls with no fish are counted as having 1 species
  mutate(no_species = case_when(no_fish == 0 ~ 0,
                                TRUE ~ as.numeric(no_species))) %>% 
  ungroup()

```

## Number of fish 
```{r number of fish in haul}
#using paired data
data_for_testing <- trap_hauls_paired %>% 
  mutate(log_no_fish = log(no_fish+1))

#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = data_for_testing, variable = 'log_no_fish', group = 'Exp_or_Cont', paired_Tor_F = TRUE)
  
#also do non-parametric test as data is not very normal
data_for_testing %>%
  wilcox.test(data = ., no_fish ~ Exp_or_Cont, paired = TRUE)
```

## Number of fish, excluding zero hauls
```{r number of fish in haul excluding zero hauls}
#using paired data without zero hauls
data_for_testing <- trap_hauls_paired_no_zeros %>% 
  mutate(log_no_fish = log(no_fish))

#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = data_for_testing, variable = 'log_no_fish', group = 'Exp_or_Cont', paired_Tor_F = TRUE)
  
#also do non-parametric test as data is not very normal
data_for_testing %>%
  wilcox.test(data = ., no_fish ~ Exp_or_Cont, paired = TRUE)
```

## Mean length of individual fish
```{r mean fish length}
#using paired data
data_for_testing <- traps_paired %>%  
  mutate(log_length = log(Length_cm+1))

#call function to get diagnostic plots and t.test
#not paired data as individual samples (fish measurements are not paired)
testing_2sample_sig_diff(data = data_for_testing, variable = 'log_length', group = 'Exp_or_Cont', paired_Tor_F = FALSE)
  
#also do non-parametric test
data_for_testing %>%
  wilcox.test(data = ., Length_cm ~ Exp_or_Cont, paired = FALSE)
```

## Mean length of individual fish, zero hauls excluded

```{r mean fish length no zeros}
#using paired data
data_for_testing <- traps_paired_no_zeros %>% 
  mutate(log_length = log(Length_cm))

#call function to get diagnostic plots and t.test
#not paired data as individual samples (fish measurements are not paired)
testing_2sample_sig_diff(data = data_for_testing, variable = 'log_length', group = 'Exp_or_Cont', paired_Tor_F = FALSE)
  
#also do non-parametric test
data_for_testing %>%
  wilcox.test(data = ., Length_cm ~ Exp_or_Cont, paired = FALSE)
```

## Mean length of fish per haul

```{r mean length of fish per haul}
#using only paired data
data_for_testing <- trap_hauls_paired %>% 
  mutate(log_length = log(mean_length_cm+1))

#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = data_for_testing, variable = 'log_length', group = 'Exp_or_Cont', paired_Tor_F = TRUE)
  
#also do non-parametric test for double check
data_for_testing %>%  
  wilcox.test(data = ., mean_length_cm ~ Exp_or_Cont, paired = TRUE)
```

## Mean length of fish per haul, zeros excluded

```{r mean length of fish per haul, zeros excluded}
#using only paired data
data_for_testing <- trap_hauls_paired_no_zeros %>% 
  mutate(log_length = log(mean_length_cm+1))

#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = data_for_testing, variable = 'log_length', group = 'Exp_or_Cont', paired_Tor_F = TRUE)
  
#also do non-parametric test for double check
data_for_testing %>%  
  wilcox.test(data = ., mean_length_cm ~ Exp_or_Cont, paired = TRUE)
```

## Mean weight of individual fish

```{r mean weight of individual fish}
#using all data
data_for_testing <- traps_paired %>% 
  mutate(log_weight = log(Weight_Grams+1))

#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = data_for_testing, variable = 'Weight_Grams', group = 'Exp_or_Cont', paired_Tor_F = FALSE)
testing_2sample_sig_diff(data = data_for_testing, variable = 'log_weight', group = 'Exp_or_Cont', paired_Tor_F = FALSE)
  
#also do non-parametric test for double check
data_for_testing %>%  
  wilcox.test(data = ., Weight_Grams ~ Exp_or_Cont, paired = FALSE)
```

## Mean weight of individual fish, zeros excluded

```{r mean weight of individual fish, zeros excluded}
#using all data
data_for_testing <- traps_paired_no_zeros %>% 
  mutate(log_weight = log(Weight_Grams))

#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = data_for_testing, variable = 'Weight_Grams', group = 'Exp_or_Cont', paired_Tor_F = FALSE)
testing_2sample_sig_diff(data = data_for_testing, variable = 'log_weight', group = 'Exp_or_Cont', paired_Tor_F = FALSE)
  
#also do non-parametric test for double check
data_for_testing %>%  
  wilcox.test(data = ., Weight_Grams ~ Exp_or_Cont, paired = FALSE)
```

## Mean total biomass of fish in haul

```{r mean total biomass of fish in haul}
#using all data
data_for_testing <- trap_hauls_paired %>% 
  mutate(log_weight = log(total_biomass_kg+1))

#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = data_for_testing, variable = 'total_biomass_kg', group = 'Exp_or_Cont', paired_Tor_F = TRUE)
testing_2sample_sig_diff(data = data_for_testing, variable = 'log_weight', group = 'Exp_or_Cont', paired_Tor_F = TRUE)
  
#also do non-parametric test for double check
data_for_testing %>%  
  wilcox.test(data = ., total_biomass_kg ~ Exp_or_Cont, paired = TRUE)
```

## Mean total biomass of fish in haul, zeros excluded

```{r mean total biomass of fish in haul, zeros excluded}
#using all data
data_for_testing <- trap_hauls_paired_no_zeros %>% 
  mutate(log_weight = log(total_biomass_kg))

#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = data_for_testing, variable = 'total_biomass_kg', group = 'Exp_or_Cont', paired_Tor_F = TRUE)
testing_2sample_sig_diff(data = data_for_testing, variable = 'log_weight', group = 'Exp_or_Cont', paired_Tor_F = TRUE)
  
#also do non-parametric test for double check
data_for_testing %>%  
  wilcox.test(data = ., total_biomass_kg ~ Exp_or_Cont, paired = TRUE)
```

## Mean number of species per haul

```{r no. species per haul}
#using all data
data_for_testing <- trap_hauls_paired %>% 
  mutate(log_no_species = log(no_species+1))

#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = data_for_testing, variable = 'no_species', group = 'Exp_or_Cont', paired_Tor_F = TRUE)
testing_2sample_sig_diff(data = data_for_testing, variable = 'log_no_species', group = 'Exp_or_Cont', paired_Tor_F = TRUE)

#also do non-parametric test for double check
data_for_testing %>%  
  wilcox.test(data = ., no_species ~ Exp_or_Cont, paired = TRUE)
```

## Mean number of species per haul, zeros excluded

```{r no. species per haul, zeros excluded}
#using all data
data_for_testing <- trap_hauls_paired_no_zeros %>% 
  mutate(log_no_species = log(no_species))

#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = data_for_testing, variable = 'no_species', group = 'Exp_or_Cont', paired_Tor_F = TRUE)
testing_2sample_sig_diff(data = data_for_testing, variable = 'log_no_species', group = 'Exp_or_Cont', paired_Tor_F = TRUE)

#also do non-parametric test for double check
data_for_testing %>%  
  wilcox.test(data = ., no_species ~ Exp_or_Cont, paired = TRUE)
```

## Compare species diversity

```{r compare species diversity}

#get species abundance matrix - using paired data only
species_count_matrix <- traps_paired %>% 
  group_by(TrapID, Date_YMD, Exp_or_Cont, common, days_since_last_haul, design, location_exposure) %>% 
  count(common) %>% 
  ungroup() %>%
  #this creates a HaulID using TrapID and date
	unite(TrapID,Date_YMD,col="HaulID",sep="_") %>%
  spread(common,n,fill=0) 

#calculate diversity for each haul
diversity_matrix <- species_count_matrix %>% 
  select(-HaulID, -Exp_or_Cont, -days_since_last_haul, -design, -location_exposure) %>% 
  diversity() %>% 
  as_tibble()

#re-attach metadata
trap_diversity <- species_count_matrix %>% 
  select(HaulID, Exp_or_Cont, days_since_last_haul, design, location_exposure) %>% 
  bind_cols(diversity_matrix)

#boxplot of Shannon index
trap_diversity %>% 
  ggplot(aes(x = Exp_or_Cont, y= value, colour = Exp_or_Cont)) +
  geom_boxplot() +
  geom_point() +
  labs(x= "", y = "Shannon diversity index")

#plot pairs
trap_diversity %>% 
  mutate(HaulID_mod = str_sub(HaulID, start = 2)) %>% 
  ggplot(aes(x=Exp_or_Cont, y = value, colour = Exp_or_Cont)) +
  geom_point() +
  geom_line(aes(group = HaulID_mod))

#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = trap_diversity, variable = 'value', group = 'Exp_or_Cont', paired_Tor_F = TRUE)

trap_diversity %>% 
  wilcox.test(data = ., value ~ Exp_or_Cont, paired = TRUE)
```

## Alternative species diversity comparison
```{r compare effective number of species}
#calculate effective number of species by transforming Shannon index (see http://www.loujost.com/Statistics%20and%20Physics/Diversity%20and%20Similarity/How%20to%20compare%20the%20diversities%20of%20two%20communities.htm)

trap_diversity <- trap_diversity %>% 
  mutate(eff_diversity = exp(value))

#website and associated paper (http://www.loujost.com/Statistics%20and%20Physics/Diversity%20and%20Similarity/JostEntropy%20AndDiversity.pdf) says that it is better to compare effective diversity, so try this method too. Note that comparing paired diversity indices also seems to be a justifiable method judging by some researchgate conversations.

#Boxplot of effective diversity
trap_diversity %>% 
  ggplot(aes(x = Exp_or_Cont, y= eff_diversity, colour = Exp_or_Cont)) +
  geom_boxplot() +
  geom_point() +
  labs(x= "", y = "Effective diversity")

#plot pairs
trap_diversity %>% 
  mutate(HaulID_mod = str_sub(HaulID, start = 2)) %>% 
  ggplot(aes(x=Exp_or_Cont, y = eff_diversity, colour = Exp_or_Cont)) +
  geom_point() +
  geom_line(aes(group = HaulID_mod))
  
#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = trap_diversity, variable = 'eff_diversity', group = 'Exp_or_Cont', paired_Tor_F = TRUE)

trap_diversity %>% 
  wilcox.test(data = ., eff_diversity ~ Exp_or_Cont, paired = TRUE)
```

## Compare speies diversity, zeros excluded

```{r compare species diversity, zeros excluded}

#get species abundance matrix - using paired data, zeris excluded
species_count_matrix <- traps_paired_no_zeros %>% 
  group_by(TrapID, Date_YMD, Exp_or_Cont, common, days_since_last_haul, design, location_exposure) %>% 
  count(common) %>% 
  ungroup() %>%
  #this creates a HaulID using TrapID and date
	unite(TrapID,Date_YMD,col="HaulID",sep="_") %>%
  spread(common,n,fill=0) 

#calculate diversity for each haul
diversity_matrix <- species_count_matrix %>% 
  select(-HaulID, -Exp_or_Cont, -days_since_last_haul, -design, -location_exposure) %>% 
  diversity() %>% 
  as_tibble()

#re-attach metadata
trap_diversity <- species_count_matrix %>% 
  select(HaulID, Exp_or_Cont, days_since_last_haul, design, location_exposure) %>% 
  bind_cols(diversity_matrix)

#boxplot of Shannon index
trap_diversity %>% 
  ggplot(aes(x = Exp_or_Cont, y= value, colour = Exp_or_Cont)) +
  geom_boxplot() +
  geom_point() +
  labs(x= "", y = "Shannon diversity index")

#plot pairs
trap_diversity %>% 
  mutate(HaulID_mod = str_sub(HaulID, start = 2)) %>% 
  ggplot(aes(x=Exp_or_Cont, y = value, colour = Exp_or_Cont)) +
  geom_point() +
  geom_line(aes(group = HaulID_mod))

#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = trap_diversity, variable = 'value', group = 'Exp_or_Cont', paired_Tor_F = TRUE)

trap_diversity %>% 
  wilcox.test(data = ., value ~ Exp_or_Cont, paired = TRUE)
```

## Alternative species diversity comparison, zeros excluded
```{r compare effective number of species, zeros excluded}
#calculate effective number of species by transforming Shannon index (see http://www.loujost.com/Statistics%20and%20Physics/Diversity%20and%20Similarity/How%20to%20compare%20the%20diversities%20of%20two%20communities.htm)

trap_diversity <- trap_diversity %>% 
  mutate(eff_diversity = exp(value))

#website and associated paper (http://www.loujost.com/Statistics%20and%20Physics/Diversity%20and%20Similarity/JostEntropy%20AndDiversity.pdf) says that it is better to compare effective diversity, so try this method too. Note that comparing paired diversity indices also seems to be a justifiable method judging by some researchgate conversations.

#Boxplot of effective diversity
trap_diversity %>% 
  ggplot(aes(x = Exp_or_Cont, y= eff_diversity, colour = Exp_or_Cont)) +
  geom_boxplot() +
  geom_point() +
  labs(x= "", y = "Effective diversity")

#plot pairs
trap_diversity %>% 
  mutate(HaulID_mod = str_sub(HaulID, start = 2)) %>% 
  ggplot(aes(x=Exp_or_Cont, y = eff_diversity, colour = Exp_or_Cont)) +
  geom_point() +
  geom_line(aes(group = HaulID_mod))
  
#call function to get diagnostic plots and t.test
testing_2sample_sig_diff(data = trap_diversity, variable = 'eff_diversity', group = 'Exp_or_Cont', paired_Tor_F = TRUE)

trap_diversity %>% 
  wilcox.test(data = ., eff_diversity ~ Exp_or_Cont, paired = TRUE)
```