---
title: "Length-frequency"
author: "Jason Flower"
date: "23/06/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(knitr)
library(paletteer)
library(cowplot)
library(gt)

source("../functions/ks_test_ecdf_generic.R") #for Kolmogorov-Smirnov tests and ECDF plots
source("../functions/ks_test_ecdf.R") #for species K-S tests
source("../functions/hist_plot_func_lmat.R") #for length-frequency plots with lines for length at maturity and mean length

plot_colours <- c(Control = "#ca0020", Experimental = "#92c5de")
```

```{r data import and tidying, message=FALSE}
#full dataset (each fish)
traps <- read_csv("../data/traps_data.csv", col_types = cols_only(
  TrapID = col_character(),
  Surveyors = col_character(),
  Length_cm = col_double(),
  LengthType_FL_TL = col_character(),
  Weight_Grams = col_double(),
  sciname = col_character(),
  Date_YMD = col_date(format = ""),
  Exp_or_Cont = col_character(),
  design = col_character(),
  Deep_Shallow = col_character(),
  days_since_last_haul = col_double(),
  common = col_character(),
  family = col_character(),
  N_coord = col_double(),
  W_coord = col_double(),
  location_exposure = col_character(),
  depth_m = col_double(),
  zero_or_1_fish = col_double()
))

#same dataset, zero fish hauls excluded
traps_no_zeros <- traps %>% 
  filter(zero_or_1_fish != 0)

#to get only paired trap data, summarise to get mean length for each trap, spread the data, and then use na.omit to drop any traps that don't have a mean length value for both Control and Experimental. Gather the data back together
traps_paired <- traps %>% 
  group_by(Date_YMD, TrapID, Exp_or_Cont) %>%
  summarise(Length_mean = mean(Length_cm)) %>%
  #filter(Length_mean != 0) %>%
  mutate(TrapNo = str_extract(TrapID, "\\d+")) %>%
  ungroup() %>%
  select(-TrapID) %>% #need to remove this otherwise spread doesn't work as needed
  spread(Exp_or_Cont, Length_mean) %>%
  na.omit() %>%
  gather(key = "Exp_or_Cont", value = "Length_mean", c("Control", "Experimental")) %>%
  #rebuild TrapID code
  mutate(TrapID = str_c(str_sub(Exp_or_Cont, 1,1), TrapNo)) %>%
  select(-TrapNo) %>%
  distinct(TrapID, Date_YMD) %>%
  inner_join(traps)

#make a dataset excluding the traps hauls with no fish
traps_paired_no_zeros <- traps %>% 
  filter(common != "No fish") %>%
  group_by(Date_YMD, TrapID, Exp_or_Cont) %>%
  summarise(Length_mean = mean(Length_cm)) %>%
  #filter(Length_mean != 0) %>%
  mutate(TrapNo = str_extract(TrapID, "\\d+")) %>%
  ungroup() %>%
  select(-TrapID) %>% #need to remove this otherwise spread doesn't work as needed
  spread(Exp_or_Cont, Length_mean) %>%
  na.omit() %>%
  gather(key = "Exp_or_Cont", value = "Length_mean", c("Control", "Experimental")) %>%
  #rebuild TrapID code
  mutate(TrapID = str_c(str_sub(Exp_or_Cont, 1,1), TrapNo)) %>%
  select(-TrapNo) %>%
  distinct(TrapID, Date_YMD) %>%
  inner_join(traps)

#import data summarized at the haul level
trap_hauls <- read_csv("../data/traps_hauls.csv", col_types = cols_only(
  TrapID = col_character(),
  Date_YMD = col_date(format = ""),
  Exp_or_Cont = col_character(),
  design = col_character(),
  days_since_last_haul = col_double(),
  depth_m = col_double(),
  N_coord = col_double(),
  W_coord = col_double(),
  location_exposure = col_character(),
  mean_length_cm = col_double(),
  total_biomass_kg = col_double(),
  no_fish = col_double(),
  no_species = col_double()
))

#summarize data for hauls
trap_hauls_no_zero <- traps_no_zeros %>% 
  group_by(TrapID, Date_YMD, Exp_or_Cont) %>% 
  summarise(mean_length_cm = mean(Length_cm), total_biomass_kg = sum(Weight_Grams)/1000, no_fish = sum(zero_or_1_fish), no_species = n_distinct(sciname), depth_m = mean(depth_m), days_since_last_haul = mean(days_since_last_haul)) %>%
  mutate(no_species = case_when(no_fish == 0 ~ 0,
                                TRUE ~ as.numeric(no_species))) %>% 
  ungroup()

trap_hauls_paired <- traps_paired %>% 
  group_by(TrapID, Date_YMD, Exp_or_Cont) %>% 
  summarise(mean_length_cm = mean(Length_cm), total_biomass_kg = sum(Weight_Grams)/1000, no_fish = sum(zero_or_1_fish), no_species = n_distinct(sciname), depth_m = mean(depth_m), days_since_last_haul = mean(days_since_last_haul)) %>%
  #need this otherwise traps hauls with no fish are counted as having 1 species
  mutate(no_species = case_when(no_fish == 0 ~ 0,
                                TRUE ~ as.numeric(no_species))) %>% 
  ungroup()

trap_hauls_paired_no_zeros <- traps_paired_no_zeros %>% 
  group_by(TrapID, Date_YMD, Exp_or_Cont) %>% 
  summarise(mean_length_cm = mean(Length_cm), total_biomass_kg = sum(Weight_Grams)/1000, no_fish = sum(zero_or_1_fish), no_species = n_distinct(sciname), depth_m = mean(depth_m), days_since_last_haul = mean(days_since_last_haul)) %>%
  #need this otherwise traps hauls with no fish are counted as having 1 species
  mutate(no_species = case_when(no_fish == 0 ~ 0,
                                TRUE ~ as.numeric(no_species))) %>% 
  ungroup()
```

## Length-frequency histogram, paired trap hauls excluding zeros
```{r length-frequency histogram}

summary_stats <- traps_paired_no_zeros %>% 
  group_by(Exp_or_Cont) %>% 
  summarise(mean_length = mean(Length_cm))

traps_paired_no_zeros %>% 
  filter(Length_cm<50) %>% 
  ggplot(aes(x=Length_cm, fill= Exp_or_Cont)) +
    geom_histogram(binwidth = 1) +    
    geom_vline(data = summary_stats, aes(xintercept = mean_length), colour = "black", linetype = "dashed", show.legend = FALSE) +
    theme_bw() +
    guides(fill = FALSE) +
    theme(axis.text = element_text(size = 10), axis.title = element_blank(), strip.text.x = element_text(size = 10)) +
    facet_wrap(~Exp_or_Cont, nrow = 2) +
    scale_fill_manual(values = plot_colours)

ggsave("../outputs/figures/length_freq_paired_simple.png", width = 16, height = 8)

ggsave("../outputs/figures/Fig 4.tiff", dpi = 300, width = 19, height = 12, units = "cm", compression = "lzw")
```

## Length-frequency histogram, with species coloured and % above bars
```{r length-frequency histogram with species and percent}
#species factored data for plotting
species_fct <- traps_paired_no_zeros %>% 
  mutate(species_lumped = fct_lump(common, prop = 0.02)) %>%
  filter(Length_cm<50)

#control traps histogram
control_hist_species_perc <- traps_paired_no_zeros %>% 
  filter(Exp_or_Cont == "Control") %>% 
  filter(Length_cm<50) %>% 
  ggplot((aes(x=Length_cm))) +
  stat_bin(breaks = seq(9,50,1), geom = "text", aes(label=sprintf("%0.1f",100*..count../sum(..count..))), vjust = -0.5, size = 2) +
  geom_histogram(data = filter(species_fct, Exp_or_Cont == "Control"), aes(x=Length_cm, fill = species_lumped), breaks = seq(9,50,1)) +
  scale_fill_paletteer_d("ggsci::category20_d3") +
  theme_bw() +
  theme(axis.text = element_text(size = 10), axis.title = element_text(size = 12), legend.title = element_text(size = 10), legend.text = element_text(size = 8)) +
  labs(x = NULL, y = "Number of fish", fill = "Species") 

#experimental traps histogram
exp_hist_species_perc <- traps_paired_no_zeros %>% 
  filter(Exp_or_Cont == "Experimental") %>% 
  filter(Length_cm<50) %>% 
  ggplot((aes(x=Length_cm))) +
  stat_bin(breaks = seq(9,50,1), geom = "text", aes(label=sprintf("%0.1f",100*..count../sum(..count..))), vjust = -0.5, size = 2) +
  geom_histogram(data = filter(species_fct, Exp_or_Cont == "Experimental"), aes(x=Length_cm, fill = species_lumped), breaks = seq(9,50,1)) +
  scale_fill_paletteer_d("ggsci::category20_d3") +
  theme_bw() +
  theme(axis.text = element_text(size = 10), axis.title = element_text(size = 12), legend.title = element_text(size = 10), legend.text = element_text(size = 8)) +
  labs(x = "Length (cm)", y = "Number of fish", fill = "Species")

# extract the legend from one of the plots and create some space to the left of the legend
legend <- get_legend(control_hist_species_perc + theme(legend.box.margin = margin(0, 0, 0, 12)))

species_hist_plots <- plot_grid(control_hist_species_perc + theme(legend.position = "none"),
          exp_hist_species_perc + theme(legend.position = "none"),
          nrow = 2,
          labels = c("a", "b"),
          label_size = 12)

plot_grid(species_hist_plots, legend, rel_widths = c(2, .55))

ggsave("../outputs/figures/length_freq_species_perc.png", width = 16, height = 10)
ggsave("../outputs/figures/S2 Fig.tiff", dpi = 300, width = 19, height = 14, units = "cm", compression = "lzw")
```

## % of total number of fish in each 1cm size class
```{r % fish in traps}
#table of relative % of fish in size classes
kable(traps_paired_no_zeros %>% 
  mutate(binned_length = cut(Length_cm, breaks = seq(0,max(Length_cm),1))) %>% 
  count(Exp_or_Cont, binned_length) %>% 
  group_by(Exp_or_Cont) %>% 
  mutate(perc_total = 100*n/sum(n)) %>% 
  pivot_wider(names_from = Exp_or_Cont, values_from = c(perc_total, n), values_fill = list(perc_total = 0, n = 0)) %>% 
  arrange(binned_length) %>% 
  mutate(cum_perc_Control = cumsum(perc_total_Control), cum_perc_Exp = cumsum(perc_total_Experimental)),
  col.names = c("Binned length (cm)", "% of total number of fish in control", "% of total number of fish in experimental", "Number in control", "Number in experimental", "Cumulative % total in control", "Cumulative % total in experimental"),
  caption = "Table of relative % of fish in control and experimental traps from each 1 cm size class")
```

## Kolmogorov-Smirnov test for all paired hauls excluding zeros
```{r K-S tests}
#all paired traps excluding zeros
ks_testing_generic(traps_paired_no_zeros)
```

## Kolmogorov-Smirnov test for all paired hauls excluding zeros, fish less than 20cm length
```{r K-S tests less than 20cm}
#all paired traps excluding zeros, fish less than 20 cm length
traps_paired_no_zeros %>% 
  filter(Length_cm <20) %>% 
  ks_testing_generic()
```

## Kolmogorov-Smirnov test for all paired hauls excluding zeros, fish less than 15 cm length
```{r K-S tests less than 15cm}
#all paired traps excluding zeros, fish less than 15 cm length
traps_paired_no_zeros %>% 
  filter(Length_cm <15) %>% 
  ks_testing_generic()
```

## Species Kolmogorov-Smirnov tests
```{r ks tests species}
#Kolmogorov-Smirnov tests to see if length-frequency distributions of species are the same. Using only species with more than 10 samples in both control and experimental traps - see Razali and Wah 2011 for discussion of low power of tests at small sample sizes
#Also plotting empirical cumulative distribution function and D-value which is used to calculate K-S test significance
  
#get list of species that have at least 10 individuals in both control and experimental traps
species_more_than_10_both_traps <- traps %>% 
  filter(common != "No fish") %>%
  group_by(common, Exp_or_Cont) %>% 
  summarize(n=n()) %>%
  filter(n>10) %>% 
  ungroup %>% 
  spread(Exp_or_Cont, n) %>% 
  na.omit() %>% 
  gather(key = "Exp_or_Cont", value = "n", c("Control", "Experimental")) %>% 
  distinct(common) %>% 
  pull()

ecdf_hist_plots<- traps %>% 
  filter(common %in% species_more_than_10_both_traps) %>% 
  #this splits the data into multiple dataframes corresponding to each species
  split(.$common) %>% 
  map(~ks_testing(data = .))

iwalk(ecdf_hist_plots, ~{
     pdf(paste0("../outputs/figures/species_ecdf_hist/", .y, ".pdf") )
     print(.x)
     dev.off()
})
```

## Length-frequency plots for short, medium and long soak times
```{r l-f soak time}
soak_time_lengt_freq_allplots <- traps_no_zeros %>% 
  filter(Length_cm < 50) %>% 
  mutate(soak_time_fct = case_when(days_since_last_haul<8 ~ "short",
                                   days_since_last_haul>8 & days_since_last_haul<15 ~ "medium",
                                   days_since_last_haul >15 ~ "long")) %>%
  mutate(soak_time_fct = as.factor(soak_time_fct),
         soak_time_fct = fct_relevel(soak_time_fct, c("short", "medium", "long"))) %>% 
  mutate(species_lumped = fct_lump(common, prop = 0.02)) %>%
  group_split(soak_time_fct) %>% 
  map(~ggplot(., aes(x=Length_cm, fill = species_lumped)) +
  geom_histogram(breaks = seq(9,50,1)) +
  scale_fill_paletteer_d("ggsci::category20_d3") +
  #discrete_scale("fill", "manual", colorRampPalette(brewer.pal(9, "Set1")),  guide = guide_legend(ncol = 1, title.position = "top")) +
  labs(x = "Length (cm)", y = "Number of fish") +
  theme_bw() +
  theme(axis.text = element_text(size = 10), axis.title = element_text(size = 12), legend.title = element_text(size = 10), legend.text = element_text(size = 10)))

#for combined legend with 3 plots (from: https://wilkelab.org/cowplot/articles/shared_legends.html)
#note I need to definte the package cowplot:: because sjplot also has plot_grid but must work slightly differently

legend <- get_legend(
  # create some space to the left of the legend
  soak_time_lengt_freq_allplots[[1]] + guides(fill=guide_legend(title="Species"))
)

combined_plot <- cowplot::plot_grid(soak_time_lengt_freq_allplots[[1]] + theme(legend.position="none", axis.title.x = element_blank(), axis.title.y = element_blank()),
                   soak_time_lengt_freq_allplots[[2]] + theme(legend.position="none", axis.title.x = element_blank(), axis.title.y = element_blank()),
                   soak_time_lengt_freq_allplots[[3]] + theme(legend.position="none", axis.title.y = element_blank()),
                   align = "v",
                   ncol = 1,
                   scale = 0.92,
                   labels = "auto")  

cowplot::plot_grid(combined_plot, legend, rel_widths = c(1,0.3)) +
  draw_label("Number of fish", x= 0, y=0.5, vjust= 1.5, angle=90, size = 12)

ggsave("../outputs/figures/soak_time_length_freq_species.png", width = 16, height = 12)
ggsave("../outputs/figures/S5 Fig.tiff", dpi = 300, width = 19, height = 12, units = "cm", compression = "lzw")

#get number of trap hauls in each soak time grouping
traps_no_zeros %>% 
  mutate(soak_time_fct = case_when(days_since_last_haul<8 ~ "short",
                                   days_since_last_haul>8 & days_since_last_haul<15 ~ "medium",
                                   days_since_last_haul >15 ~ "long")) %>% 
  mutate(soak_time_fct = as.factor(soak_time_fct)) %>% 
  group_by(TrapID, Date_YMD, Exp_or_Cont) %>% 
  count(soak_time_fct) %>% 
  group_by(soak_time_fct) %>% 
  tally()

#mean fish length for long, medium and short soak times
traps_no_zeros %>% 
  mutate(soak_time_fct = case_when(days_since_last_haul<8 ~ "short",
                                   days_since_last_haul>8 & days_since_last_haul<15 ~ "medium",
                                   days_since_last_haul >15 ~ "long")) %>% 
  mutate(soak_time_fct = as.factor(soak_time_fct)) %>% 
  group_by(soak_time_fct) %>% 
  summarise(mean_length = mean(Length_cm))

#change in % of species total with incresing soak time - table

traps_no_zeros %>% 
  filter(Length_cm < 50) %>% 
  mutate(soak_time_fct = case_when(days_since_last_haul<8 ~ "short",
                                   days_since_last_haul>8 & days_since_last_haul<15 ~ "medium",
                                   days_since_last_haul >15 ~ "long")) %>%
  mutate(soak_time_fct = as.factor(soak_time_fct),
         soak_time_fct = fct_relevel(soak_time_fct, c("short", "medium", "long"))) %>% 
  count(soak_time_fct, common) %>% 
  pivot_wider(names_from = soak_time_fct, values_from = n, values_fill = list(n=0)) %>% 
  mutate(prop_fish_short = 100*short/sum(short), prop_fish_medium = 100*medium/sum(medium), prop_fish_long = 100*long/sum(long)) %>% 
  arrange(desc(prop_fish_short))
```

## Length-frequency plots for species with more than 10 individuals in control and experimental plots
Using Fishbase Lmat values. I will not use this in results because Fishbase Lmat values are not very reliable: sometimes based on very small sample sizes and not from islands close to Montserrat

```{r length frequency with Lmat}
scinames_more_than_10_both_traps<- traps_paired_no_zeros %>% 
  filter(common %in% species_more_than_10_both_traps) %>% 
  distinct(sciname, common) %>% 
  pull(sciname)
  
#pull Lmat values from fishbase
lmat_fishbase <- rfishbase::maturity(scinames_more_than_10_both_traps, fields = c("Species", "Lm"))

#filter out NAs and use mean for cases where there are multiple Lmat values
lmat_fishbase <- lmat_fishbase %>% 
  filter(!is.na(Lm)) %>% 
  group_by(Species) %>% 
  summarise(Lmat = mean(Lm))

lmat_plot <- traps_paired_no_zeros %>% 
  inner_join(., lmat_fishbase, by = c("sciname" = "Species")) %>% 
  #this splits the data into multiple dataframes corresponding to each species
  split(.$common) %>% 
  map(~hist_plot_func_lmat(data = ., x = 'Length_cm', group = 'Exp_or_Cont', bin_widths=1, Lmat = 'Lmat'))

pdf("../outputs/figures/Lmat-plots/lmat_plots.pdf")
lmat_plot
dev.off()

range(lmat_fishbase$Lmat)
#gridExtra::grid.arrange(grobs = lmat_plot, ncol = 2)

```

## Length-frequency plots for species with more than 10 individuals in control and experimental plots
Using local Lmat values

```{r length-frequency with Lmat local values}

#view list of species I need Lmat values for
sort(scinames_more_than_10_both_traps)

#species are "Banded butterflyfish"  "Black margate"         "Blue tang"             "Bluestriped grunt"     "Doctorfish"           
# [6] "French grunt"          "Honeycomb cowfish"     "Lane snapper"          "Pluma porgy"           "Queen triggerfish"    
#[11] "Red hind"              "Rock beauty"           "Schoolmaster"          "Smooth trunkfish"      "Spotfin butterflyfish"
#[16] "Squirrelfish"          "Whitespotted filefish" "Yellow goatfish" 

#match Lmat values from Montserrat Fisheries Assessment 2016 report (SFG 2016 reference), these are from following csv: https://github.com/lennon-thomas/Montserrat-Fisheries-Assessment/blob/239b3157abc7029b4c141d13fea823a1d5565955/Length_Assessment/MNI_LH_spread.csv
sfg_2016 <- read_csv("../data/Montserrat_LH_parameters_SFG_2016.csv")

lmat_values <- tibble(common = species_more_than_10_both_traps) %>% 
  left_join(., select(sfg_2016, CommName, Mat50, Mat95), by = c("common" = "CommName"))

#use values available in Ault, J.S., Smith, S.G., Luo, J., Monaco, M.E. & Appeldoorn, R.S. (2008). Length-based assessment of sustainability benchmarks for coral reef fishes in Puerto Rico. Environmental Conservation, 35, 221. https://doi.org/10.1017/S03768929080050437

lmat_values <- lmat_values %>% 
  mutate(Mat50 = case_when(common == "Bluestriped grunt" ~ 20.5,
                           common == "French grunt" ~ 17.7,
                           common == "Lane snapper" ~ 20.6,
                           common == "Schoolmaster" ~ 14.5,
                           TRUE~as.numeric(Mat50)))

lmat_plot_local <- traps_paired_no_zeros %>% 
  inner_join(., lmat_values, by = "common") %>% 
  filter(!is.na(Mat50)) %>% 
  #this splits the data into multiple dataframes corresponding to each species
  split(.$common) %>% 
  map(~hist_plot_func_lmat(data = ., x = 'Length_cm', group = 'Exp_or_Cont', bin_widths=1, Lmat = 'Mat50'))

pdf("../outputs/figures/Lmat-plots/lmat_plots_local.pdf")
lmat_plot_local
dev.off()

```

## Table of Lmat and mean length 
```{r Lmat mean length table}

mean_length_Mat50_table <- traps_paired_no_zeros %>% 
  select(Exp_or_Cont, common, sciname, Length_cm) %>% 
  inner_join(lmat_values, by = "common") %>% 
  filter(!is.na(Mat50)) %>% 
  group_by(Exp_or_Cont, common, sciname) %>% 
  summarise(mean_length = mean(Length_cm),
            Mat50 = mean(Mat50)) %>%
  ungroup() %>% 
  arrange(common, Exp_or_Cont) %>%
  pivot_wider(names_from = Exp_or_Cont, values_from = mean_length) %>% 
  mutate("Mean length greater than Mat50?" = case_when(Control > Mat50 | Experimental>Mat50 ~ TRUE,
                                                       Control < Mat50 | Experimental<Mat50 ~ FALSE))

mean_length_Mat50_table %>% 
  mutate(species = paste0(common, " (", sciname,")")) %>% 
  select(-common, -sciname) %>%
  relocate(species, .before = Mat50) %>% 
  mutate(Mat50 = sprintf("%.1f", Mat50),
         Control = sprintf("%.1f", Control),
         Experimental = sprintf("%.1f", Experimental)) 

```
## Proportions of fish below Lmat
```{r proportions of fish below  Lmat}
perc_less_Lmat <- traps_paired_no_zeros %>% 
  select(Exp_or_Cont, common, sciname, Length_cm) %>% 
  inner_join(lmat_values, by = "common") %>% 
  filter(!is.na(Mat50)) %>% 
  mutate(Length_less_Lmat = case_when(Length_cm<Mat50 ~ 1,
                                         TRUE ~ as.numeric(0))) %>% 
  group_by(Exp_or_Cont, common, sciname) %>% 
  summarise(Mat50 = mean(Mat50),
            mean_length = mean(Length_cm),
            perc_less_Lmat = sum(Length_less_Lmat)/n()) %>% 
  arrange(common, Exp_or_Cont) %>% 
  ungroup()

perc_less_Lmat

#outputs table
  perc_less_Lmat %>% 
    select(-Mat50, -mean_length) %>% 
    pivot_wider(names_from = Exp_or_Cont, values_from = perc_less_Lmat) %>%
    mutate(perc_higher_in_control = case_when(Control > Experimental ~ "True",
                                              Control < Experimental ~ "False",
                                              Control == Experimental ~ "Equal")) %>% 
    mutate(Control = sprintf("%.1f", Control*100),
           Experimental = sprintf("%.1f", Experimental*100)) %>%
    mutate(species = paste0(common, " (", sciname,")")) %>% 
    select(-common, -sciname) %>% 
    relocate(species, .before = Control) %>% 
    arrange(desc(perc_higher_in_control), species) %>% 
    gt() %>% 
    cols_label(species = "Species name",
               Control = "Percent of fish species less than Lmat in control trap",
               Experimental = "Percent of fish species less than Lmat in experimental trap",
               perc_higher_in_control = "More immature fish in control trap?") %>% 
    gtsave("Lmat_exp_control.rtf", path = "../outputs/")


  
```


